\section{Definitional Equality}
\label{sec:equality}

Some properties cannot or need not use induction or some more
sophisticated technique, since they are true by definition. Examples
are properties for fully polymorphic functions such as this definition
of \hs{id} in the SK-calculus:

\begin{code}
s f g x = f x (g x)
k x y = x
id x = x

prop_skk_id :: Prop (a -> a)
prop_skk_id = s k k =:= id
\end{code}

The generated conjecture stated with function pointers as this:

\begin{equation*}
\app{ (\app {\ptr{s}} {\ptr{k}} )
    }{\ptr{k}} = \ptr{id}
\end{equation*}

However, it is not provable in this form. The next section will solve this.

\subsection{Extensional Equality and \texttt{seq}}

To prove the previous property we also need to have extensional
equality, postulated with this following axiom:

\begin{equation*}
\faa{f}{g} (\fa{x} \app{f}{x} = \app{g}{x}) \rightarrow f = g
\end{equation*}

\noindent
which identifies function pointers and functions composed with $@$.
One problem with extensional equality in Haskell, is that the presence
of \hs{seq} weakens it. \hs{seq} is a built in function with the
following behaviour:

\begin{code}[mathescape]
seq :: a -> b -> b
seq $\bot$ y = $\bot$
seq x y = y$, \qquad x \neq \bot$
\end{code}

It forces the first argument to weak head normal form and returns the
second. For our purposes, it is only important if the first argument
is $\bot$, then the function also returns this as it is strict in its
first argument. With \hs{seq} it is possibly to distinguish between
these two functions, breaking extensional equality since they are
otherwise observationally equal:

\begin{code}[mathescape]
f = $\bot$
g = \x -> $\bot$
\end{code}

Because \hs{seq f ()} evaluates to $\bot$, and \hs{seq g ()} evaluates
to \hs{()}, but on any argument \hs{f} and \hs{g} gets, they both
return $\bot$. Here we also need an extra axiom, which says that
anything applied to $\bot$ is $\bot$:

\begin{equation*}
\fa{x} \app{\bot}{x} = \bot
\end{equation*}

However, \hs{seq} is the only function that can tell such functions
apart, so we will ignore its presence in Haskell.  In the future,
there could be added as a switch \hs{--enable-seq}, which weakens
extensional equality appropriately.

If extensional equality is assumed we also have the property that
\hs{Prop (a -> b)} is equivalent to \hs{a -> Prop b}, by letting the
property have an extra argument that is applied to the left and right
hand side of the equality. This has two benefits. Firstly, it can
trigger other proof methods should \hs{a} or \hs{b} be concrete types
(the former for induction and the latter for approximation
lemma). Secondly, for many properties the extensionality axiom which
introduces extra steps in the proof search is not needed.

The property about \hs{s k k} is then instead translated to:

\begin{equation*}
\fa{x} \fn{s}(\ptr{k},\ptr{k},x) = \fn{id}(x)
\end{equation*}

This gives a little less unnecessary overhead to the theorem provers.

\subsection{Future Work: Concrete Concerns}
\label{sec:concreteconcerns}

This only works on non-concrete types because of the way bottoms are
added. One example of such a problem is this standard definition of
\hs{\&\&}, and a property stating its right identity:

\begin{code}
True  && a = a
False && _ = False

prop_or_right_identity :: Bool -> Prop Bool
prop_or_right_identity x = x && False =:= x
\end{code}

The translation of \hs{\&\&} makes any element in the domain that is not
the introduced constants $\fn{false}$ or $\fn{true}$ for \hs{Bool}'s
constructors, equal $\bot$. Now consider the translation of the
property:

\begin{equation*}
\fa{x} x \, \fn{\&\!\&} \, \fn{true} = x
\end{equation*}

Now this is false, take a model with another element $\diamond$ in the
domain:

$$\diamond \, \fn{\&\!\&} \, \fn{true} = \bot$$

The consequence of this is that the proof principle of definitional
equality is only used on abstract types, rather than polymorphic, as
they cannot be strict. Do not fear: the property above is trivially
proved by induction, as induction for \hs{Bool} and other non
recursive data types degenerate into mere case enumeration. However,
it can only be seen as a meta theorem as it cannot simply be added as
a lemma in the theory. This phenomena is described in more detail in
Section \ref{sec:thmlemmas}. Structural induction is described in
detail in the next section.


