\chapter{Proof Techniques}
\label{ch:proofs}

The proof developed in this thesis is called \hs{hip}, the Haskell
Inductive Prover. To use it, properties are inserted to the source
code where the definitions of the relevant functions are. As an
example, this is how the associativity of list concatenation can be
entered:

\begin{code}
import AutoPrelude

prop_app_assoc :: [a] -> [a] -> [a] -> Prop [a]
prop_app_assoc xs ys zs = xs ++ (ys ++ zs) =:= (xs ++ ys) ++ zs
\end{code}

\noindent
The arguments are universally quantified, so this property means:

\newcommand\append{+\!+}
\begin{equation*}
  \faaa{xs}{ys}{zs} xs \append (ys \append zs) = (xs \append ys) \append zs
\end{equation*}

\noindent
The equality is interpreted as equality on the constructor level: two
values are identified if they are constructed with the same
constructor and their arguments are also equal.

The infix function \hs{=:=} comes from the import, as well as the type
constructor \hs{Prop}. The type signature cannot be omitted as this is
used for some proof techniques.  Using \hs{hip} is then a matter of
saving the file, for instance to \hs{ListProps.hs}, and executing this
statement in your favourite terminal:

\begin{code}
hip ListProps.hs
\end{code}

The program will report to you which proof methods succeeded on proving
this property, if any. In this case, it is provable with all three
inductive techniques.

By importing \hs{AutoPrelude} the properties in the file are also
testable with QuickCheck, given that there are appropriate \hs{Eq} and
\hs{Arbitrary} instances provided.

The rest of this chapter explains the different proof methods
supported in this tool. Some properties are a direct consequence from
the definitions in your file. How to prove such properties is
described in Section \ref{sec:equality} about definitional
equality. The three other techniques uses induction in different
ways. Structural induction is explained in Section
\ref{sec:induction}, which uses the structure of the data types a
property quantifies over. Another method which does induction on the
recursive structure of the program is Fixed point induction,
introduced in Section \ref{sec:fixpoint}. A more subtle way of
induction is used in the Approximation Lemma, Section
\ref{sec:approx}, where the structure of the data type of the equality
is approximated.

\pagebreak

% Definitional Equality -------------------------------------------------------

\input{equality}

% Structual Induction ---------------------------------------------------------

\input{induction}

% Fixpoint Induction ----------------------------------------------------------

\input{fixpoint}

% Approximation Lemma ---------------------------------------------------------

\input{approx}
